Type-Safe Modular Parsing

### 2017-01-14 paper/paper.tex
Abstract. Introduction: halfway.

  This paper investigates solutions for the problem of \emph{modular
    parsing}. We focus on \emph{semantic} modularity and not just
  \emph{syntactic} modularity. That is, the solutions should not only
  allow complete parsers to be built out of modular parsing
  components, but also enable the parsing components to be
  \emph{modularly type-checked} and \emph{separately compiled}. We
  present a technique based parsing combinators that enables modular
  parsing. Interestingly, the modularity requirements for modular
  parsing rule out several existing parsing combinator approaches,
  which rely on some non-modular techniques. We show that Packrat
  parsing techniques, provide solutions for such modularity problems,
  and enable reasonable performance in a modular setting.
  Extensibility is achieved using \emph{multiple inheritance} and 
  \emph{Object Algebras}.  To evaluate
  the approach we conduct a case study based on the ``Types and
  Programming Languages'' interpreters. The case study the
  effectiveness at reusing parsing code from existing interpreters,
  and the total parsing code is 69\% shorter than an existing code
  based using non-modular parsing code.

>>>>

  This paper investigates solutions for the problem of \emph{modular
    parsing}. We focus on \emph{semantic} modularity and not just
  \emph{syntactic} modularity. That is, the solutions should not only
  allow complete parsers to be built out of modular parsing
  components, but also enable the parsing components to be
  \emph{modularly type-checked} and \emph{separately compiled}. We
  present a technique based on parsing combinators that enables modular
  parsing. Interestingly, the modularity requirements for modular
  parsing rule out several existing parsing combinator approaches,
  which rely on some non-modular techniques. We show that Packrat
  parsing techniques, provide solutions for such modularity problems,
  and enable reasonable performance in a modular setting.
  Extensibility is achieved using \emph{multiple inheritance} and
  \emph{Object Algebras}.  To evaluate
  the approach we conduct a case study based on the ``Types and
  Programming Languages'' interpreters. The case study shows the
  effectiveness at reusing parsing code from existing interpreters,
  and the total parsing code is 69\% shorter than an existing code
  base using non-modular parsing code.

### 2017-01-14 paper/sections/Introduction.tex
Abstract. Introduction: halfway.

To address the modularity limitations of Programming Languages several
different approaches have been proposed in the past. Existing
approaches can be broadly divided into two categories:
\emph{syntactic} or \emph{semantic} modularization
techniques. Syntactic modularization techniques are quite popular in
practice, due to their simplicity of implementation and use. 
Examples include many tools for developing Feature-Oriented Software-Product
Lines (SPLs)~\cite{Kastner11road}, some Language Workbenches~\cite{}, or extensible parser
generators~\cite{}.  Most syntactic approaches employ textual
composition techniques such as \emph{superimposition}~\cite{} to
enable the development modular program features. Such 
composition techniques collect the code for multiple features and
merge them together when a concrete combination of features is needed
for a particular program. As Kastner et. al~\cite{Kastner11road} note, 
a typical drawback of feature-oriented SPL implementations, which 
more generally applies to syntactic modularity approaches, is that
such ``\emph{implementation mechanisms lack proper
  interfaces and support neither modular type checking nor separate
  compilation}''.

@@

Semantic modularization techniques go one step further in terms of modularity,
and also enable components or features to be modularly type-checked
and separately compiled. Modular type-checking and separate
compilation are desirable properties to have from a software
engineering point-of-view. Modular type-checking can report errors 
earlier and in terms of the modular code programmers have written 
in the first place. Separate compilation avoids global compilation
steps, which can be very costly. Furthermore semantic modularization 
enables the composition of compiled binaries as well as ensuring the 
type-safety of the code composed of multiple components. Examples of semantic modularization techniques 
include various approaches to \emph{family polymorphism}~\cite{ernst01FP},
\emph{virtual classes}~\cite{Ernst:2006}, as 
well as various techniques for solving the Expression
Problem~\cite{Oliveira:2012}\bruno{more here; yanlin; Torgersen; Odersky...}.

Semantic modularization techniques are less widely used in practice
than syntactic techniques. This is partly due to the perceived need for more
sophisticated type systems, which are not available in mainstream
languages and may require more knowledge from users. However, recently,
several lightweight modularization techniques have been shown to work
in mainstream programming languages like Java or Scala. Object
Algebras~\cite{Oliveira:2012} are one such technique, which works in
Java-like languages and uses simple generics only.

So far research on semantic modularization techniques have focused on
operations that \emph{traverse} or \emph{process} extensible
datastructures, such as ASTs. Indeed many documented applications of
semantic modularization techniques focus on modularizing various
aspects of PL implementations.  However, as far as we know, there is
little work on operations that build/produce ASTs.  In particular the
problem of how to modularize parsing has not been studied in semantic
modularization approaches. At best, techniques such as NOA~\cite{Gouseti2014}
employ a syntactic modularity approach for parsing in combination with 
a semantic modularity approach for defining other language components 
that traverse or process ASTs. This is a shame because parsing is a
fundamental part of PL implementations, and it ought to made
semantically modular as well, so that the full benefits of semantic 
modularity apply.

@@

This paper presents a technique for doing semantically
modular parsing.  That is, our approach not only allows complete
parsers to be built out of modular parsing components, but also enables
those parsing components to be \emph{modularly type-checked} and
\emph{separately compiled}. Developing techniques for modular parsing 
is not without challenges. In developing our techniques we encontered 
two different classes of challenges: algorithmic challenges; and
typing/reuse challenges.

\paragraph{Algorithmic Challenges} A first challenge was todo with
  the parsing algorithms themselves. Since many parsing formalisms
  were not designed with extensibility in mind, some of the employed
  techniques were non-modular, or performed poorly in a modular
  setting.  The most widely use tools for parsing are parser
  generators, but most algorithms employed by parser generators
  require full information about the grammar to generate parsing
  code. Moreover, actions associated to grammar productions are
  typically only type-checked after the parser has been
  generated. Both problems go against our goals of semantic
  modularity.

  An alternative to parser generators are \emph{parser
    combinators}~\cite{burge1975,Wadler1985}.  At a first look, parser
  combinators seem very suitable for our purpose. Each parser
  combinator is represented by a piece of code directly in the
  programming language. Thus, in a statically typed programming
  language, such code is statically type-checked.  However many
  techniques regularly employed by parser combinators cause difficulties in a
  modular setting. In particular, many parser combinator approaches
  (including Parsec~\cite{}) routinely use \emph{left-recursion
    elimination}, \emph{priority-based matching}, and \emph{avoid
    backtracking} as much as possible. All of these are problematic in
  a modular setting as illustrated in Section~\ref{}.

  To address such algorithmic challanges, we propose a methodology 
  for implementing modular parsers built on top of an
  existing Packrat~\cite{} parsing library for Scala~\cite{}. The Packrat parsing library
  directly supports left-recursion, and a \emph{longest-match
    composition} operator. The direct support for left-recursion
  avoids the need for left-recursion elimination, which is very
  cumbersome is a modular setting. The longest-match composition 
  operator avoids the use of priorities, and it provides reasonably
  efficient optimizations for backtracking.

  \paragraph{Typing and Reusability Challenges} The second class of
  challenges were problems related to modularity, reusability and
  typing of parsing code. An immediate concern is how to extend a
  parser for an existing language or, more generally, how to compose
  parsing code for two languages. It turns out that OO mechanisms that
  provide some form of multiple inheritance,
  such as traits/mixins~\cite{scalaoverview}\bruno{cite work on traits (Ducasse)
  and mixins (Bracha)}, are very handy for this
  purpose. Essentially, traits/mixins can act as modules for the parsing code
  of different languages. This enables an approach where ASTs can be
  modelled using standard OO techniques such as the {\sc Composite} 
  pattern, while retaining the possibility of adding new language
  constructs. Section~\ref{} gives the details of this approach.

  Our ultimate goal is to allow for full extensibility: it should be
  possible to not only modularly add both new language constructs, but
  also new operations. To accomplish this goal one final tweak on our
  technique is to employ Object Algebras to allow fully extensible
  ASTs. Thus a combination of Packrat parsing, multiple inheritance 
  and Object Algebras enables a solution for semantically modular 
  parsing. Section~\ref{} gives the details of the complete approach.

@@

\item {A technique for Modular Parsing} We present a technique 
that allows the development semantically modular parsers.
The technique relies on the combination of Packrat parsing, multiple inheritance 
  and Object Algebras.

@@

All the code in the paper and the case study is written in Scala. But,
other languages that support some form of multiple inheritance
(including Java 8 with default methods~\cite{}) could in principle be
used. Our choice for Scala had todo with the concise and elegant syntax  
of the language which enable a clear presentation of our ideas.

>>>>

To address the modularity limitations of Programming Languages, several
different approaches have been proposed in the past. Existing
approaches can be broadly divided into two categories:
\emph{syntactic} or \emph{semantic} modularization
techniques. Syntactic modularization techniques are quite popular in
practice, due to their simplicity of implementation and use.
Examples include many tools for developing Feature-Oriented Software-Product
Lines (SPLs)~\cite{Kastner11road}, some Language Workbenches~\cite{}, or extensible parser
generators~\cite{}.  Most syntactic approaches employ textual
composition techniques such as \emph{superimposition}~\cite{} to
enable the development modular program features. Such
composition techniques collect the code for multiple features and
merge them together when a concrete combination of features is needed
for a particular program. As Kastner et. al~\cite{Kastner11road} note,
a typical drawback of feature-oriented SPL implementations, which
more generally applies to syntactic modularity approaches, is that
such ``\emph{implementation mechanisms lack proper
  interfaces and support neither modular type checking nor separate
  compilation}''.

@@

Semantic modularization techniques go one step further in terms of modularity,
and also enable components or features to be modularly type-checked
and separately compiled. Modular type-checking and separate
compilation are desirable properties to have from a software
engineering point-of-view. Modular type-checking can report errors
earlier and in terms of the modular code programmers have written
in the first place. Separate compilation avoids global compilation
steps, which can be very costly. Furthermore semantic modularization
enables the composition of compiled binaries as well as ensuring the
type-safety of the code composed of multiple components. Examples of semantic modularization techniques
include various approaches to \emph{family polymorphism}~\cite{ernst01FP},
\emph{virtual classes}~\cite{Ernst:2006}, as
well as various techniques for solving the Expression
Problem~\cite{Oliveira:2012,wang2016expression,torgersen2004expression,odersky2005independently}.
Semantic modularization techniques are less widely used in practice
than syntactic techniques. This is partly due to the perceived need for more
sophisticated type systems, which are not available in mainstream
languages and may require more knowledge from users. However, recently,
several lightweight modularization techniques have been shown to work
in mainstream programming languages like Java or Scala. Object
Algebras~\cite{Oliveira:2012} are one such technique, which works in
Java-like languages and uses simple generics only.

So far research on semantic modularization techniques has focused on
operations that \emph{traverse} or \emph{process} extensible
data structures, such as ASTs. Indeed many documented applications of
semantic modularization techniques focus on modularizing various
aspects of PL implementations.  However, as far as we know, there is
little work on operations that build/produce ASTs.  In particular the
problem of how to modularize parsing has not been studied in semantic
modularization approaches. At best, techniques such as NOA~\cite{Gouseti2014}
employ a syntactic modularity approach for parsing in combination with
a semantic modularity approach for defining other language components
that traverse or process ASTs. This is a shame because parsing is a
fundamental part of PL implementations, and it ought to be made
semantically modular as well, so that the full benefits of semantic
modularity apply.

@@

This paper presents a technique for doing semantically
modular parsing.  That is, our approach not only allows complete
parsers to be built out of modular parsing components, but also enables
those parsing components to be \emph{modularly type-checked} and
\emph{separately compiled}. Developing techniques for modular parsing
is not without challenges. In developing our techniques we encountered
two different classes of challenges: algorithmic challenges; and
typing/reuse challenges.

\paragraph{Algorithmic Challenges} A first challenge was to do with
  the parsing algorithms themselves. Since many parsing formalisms
  were not designed with extensibility in mind, some of the employed
  techniques were non-modular, or performed poorly in a modular
  setting.  The most widely use tools for parsing are parser
  generators, but most algorithms employed by parser generators
  require full information about the grammar to generate parsing
  code. Moreover, actions associated to grammar productions are
  typically only type-checked after the parser has been
  generated. Both problems go against our goals of semantic
  modularity.

  An alternative to parser generators are \emph{parser
    combinators}~\cite{burge1975,Wadler1985}.  At a first look, parser
  combinators seem very suitable for our purpose. Each parser
  combinator is represented by a piece of code directly in the
  programming language. Thus, in a statically typed programming
  language, such code is statically type-checked.  However many
  techniques regularly employed by parser combinators cause difficulties in a
  modular setting. In particular, many parser combinator approaches
  (including Parsec~\cite{Leijen2001}) routinely use \emph{left-recursion
    elimination}, \emph{priority-based matching}, and \emph{avoid
    backtracking} as much as possible. All of these are problematic in
  a modular setting as illustrated in Section~\ref{subsec:challenges}.

  To address such algorithmic challanges, we propose a methodology
  for implementing modular parsers built on top of an
  existing Packrat~\cite{Ford2002} parsing library for Scala~\cite{}. The Packrat parsing library
  directly supports left-recursion, and a \emph{longest-match
    composition} operator. The direct support for left-recursion
  avoids the need for left-recursion elimination, which is very
  cumbersome is a modular setting. The longest-match composition
  operator avoids the use of priorities, and it provides reasonably
  efficient optimizations for backtracking.

  \paragraph{Typing and Reusability Challenges} The second class of
  challenges was problems related to modularity, reusability and
  typing of parsing code. An immediate concern is how to extend a
  parser for an existing language or, more generally, how to compose
  parsing code for two languages. It turns out that OO mechanisms that
  provide some form of multiple inheritance,
  such as traits/mixins~\cite{scalaoverview}\bruno{cite work on traits (Ducasse)
  and mixins (Bracha)}, are very handy for this
  purpose. Essentially, traits/mixins can act as modules for the parsing code
  of different languages. This enables an approach where ASTs can be
  modelled using standard OO techniques such as the {\sc Composite}
  pattern, while retaining the possibility of adding new language
  constructs. Section~\ref{sec:inheritance} gives the details of this approach.

  Our ultimate goal is to allow for full extensibility: it should be
  possible to modularly add not only new language constructs, but
  also new operations. To accomplish this goal one final tweak on our
  technique is to employ Object Algebras to allow fully extensible
  ASTs. Thus a combination of Packrat parsing, multiple inheritance
  and Object Algebras enables a solution for semantically modular
  parsing. Section~\ref{sec:algebrasandparsing} gives the details of the complete approach.

@@

\item {{\bf A technique for Modular Parsing:}} We present a technique
that allows the development of semantically modular parsers.
The technique relies on the combination of Packrat parsing, multiple inheritance
  and Object Algebras.

@@

All the code in the paper and the case study is written in Scala. But,
other languages that support some form of multiple inheritance
(including Java 8 with default methods~\cite{}) could in principle be
used. Our choice for Scala had to do with the concise and elegant syntax
of the language which enables a clear presentation of our ideas.
